Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-02-06 content.tex
EOD

We show the usage of \verb;pirouette; on an tree generated by a non-standard 
tree model.

\begin{lstlisting}[
    language = R,
    floatplacement = ht,
    frame = single, 
    label = {lst:create_dd_tree}, 
    caption = {
      Create a Yule tree. 
      The resulting tree is shown in Figure~\ref{fig:yule_tree}.
    }
  ]
phylogeny <- create_dd_tree(n_taxa = 6, crown_age = 10)
\end{lstlisting}

by gradually introducing
its features.
First, to get an idea of the baseline error, 
we measure the Bayesian inference error when we start from a phylogeny 
generated under a known and standard tree model.
Second, to establish that the true tree prior is indeed the best, 
we also measure the inference error made by a
best candidate inference model.
Lastly, we quantify the impact of the tree prior in the Bayesian inference. 
We do so by separating the baseline error from the complete error 
running the pipeline starting from a tree generated by a non-standard tree model.

All the figures shown in this section are shown as-is, without any
aesthetical modifications. Figures showing the full workflow 
and tables showing the effective sample sized (a measure
of inference quality) can be found in the supplementary materials.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The generative and inference models are equal and standard.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\verb;pirouette; quantifies the influence of a new tree prior on BEAST2's 
inference by measuring the discrepancy between a given/true tree and a posterior 
distribution of phylogenies obtained as result of the inference process. 
Due to stochasticity, posterior trees will generally differ from the given 
phylogeny $\tau_{\mathit{G}}$, even when the tree 
prior and alignment model used for inference are the same as those 
used to generate the alignment.
Measuring this difference allows us to know the baseline error
of the \verb;pirouette; pipeline. We therefore define as 'standard tree priors' 
all the tree priors that are available 
to be used within an inference model (see Table~\ref{tab:options}).

Now we can formulate the first example research question that \verb;pirouette; 
can answer: "What is the inference error made on phylogenies created by a 
standard diversification model?"

In this example we use a standard generative tree model $\mathit{p_{G}^0}$, 
namely the Yule (pure-birth) tree model. 
We choose to use a small tree with six taxa, to keep
the calculations short and the figure more readable.
We pick a crown age of ten time units. This value is 
completely arbitrary, but it ties in with the mutation rate 
used in simulating an alignment in the next step.

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{pirouette_example_1/example_1_314/true_tree.png}
  \caption{The Yule tree, as created by Listing~\ref{lst:create_dd_tree}.}
  \label{fig:yule_tree}
\end{figure}

Experiments must be bundled in a list to work, even if only one is provided, as 
in this case:

\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single,
  label = {lst:create_gen_experiment},
  caption = {
    Create the experiments. In this case, we use only
    an experiment with the generative model,
    that will always be used in the actual inference.
  }
]
experiments <- list(generative_experiment)
\end{lstlisting}

We now have all the needed \verb;pirouette; parameters: the alignment 
parameters, the experiments and the error measure parameters.
These objects need to be bundled in a larger parameter structure, 
using \verb;create_pir_params;:

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_pir_params},
  caption = Calling \texttt{create\_pir\_params}.
]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  error_measure_params = error_measure_params
)
\end{lstlisting}

We can finally use the given Yule tree and \verb;pir_params; to measure the 
inference error made on phylogenies
created by a standard diversification model:

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_1/example_1_314/errors.png}
  \caption{
    The inference error made 
    when the generative and inference models are the same.
    The vertical dashed line indicates the median error.
  }
  \label{fig:example_1}
\end{figure}

The resulting error distribution, as shown Figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the generating tree model is standard and known.
Obtaining this control is in fact an inherent part of \verb;pirouette;. 
We call it 'twinning' and we will demonstrate it in section
\ref{Comparing to background noise} below.

@@

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model\_type} &
    \textbf{run\_if} &
    \textbf{do\_measure\_evidence} & 
    \textbf{inference model} \\ 
    \hline
    generative & always         & FALSE & JC, strict, Yule \\
    candidate  & best candidate & TRUE  & JC, strict, BD   \\
    candidate  & best candidate & TRUE  & JC, strict, CBS  \\
    ...        & ...            & ...   & ...              \\
    candidate  & best candidate & TRUE  & GTR, RLN, CCP    \\
    candidate  & best candidate & TRUE  & GTR, RLN, CEP    \\
    \hline
  \end{tabular}
  \caption{
    Inference conditions and model.
    JC: Jukes-Cantor nucleotide substitution model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
    BD: birth-death tree prior.
    GTR: GTR nucleotide substitution model.
    RLN: relaxed log-normal clock model.
    CBS: coalescent Bayesian Skyline tree prior.
    CCP: coalescent constant-population tree prior.
    CEP: coalescent exponential-population tree prior.
  }
  \label{tab:RQ2}
\end{table}

@@

So far we have measured the inference error on a tree
generated according to a known (and standard) tree diversification model. 
The goal of \verb;pirouette; is to measure the expected impact of a novel tree prior.
To do so, in this example, we will use a tree generated by a non-standard 
tree diversification model, assumed to be closely related to the Yule tree prior.
We will use the diversity-dependent tree model \cite{DDD, etienne2011diversity},
which is a birth-death model with a speciation that is dependent on the
number of species.
We measure both the baseline 
error (that serves as a control) and the full inference error, because their 
difference shows the impact of the tree model.
We obtain the baseline error by using a twin tree (see Subsection~\ref{subsec:twinning}) for 
both the generative and best candidate models.
The research question this example answers is:
"What is the inference error made from a phylogeny, 
for both a generative model and the best candidate model?"

We start from the tree generated by a diversity-dependent model,
having six taxa and a crown age of ten:
\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single, 
  label = {lst:unknown_phylogeny},
  caption = A phylogeny generated by an unknown diversification model.
]
phylogeny <- create_dd_tree(n_taxa = 6, crown_age = 10)
\end{lstlisting}
\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_3/example_3_314/true_tree.png}
  \caption{The tree derived from an unknown diversification process, 
    as created by listing~\ref{lst:unknown_phylogeny}.
    The scale overlaps with the lower branch, which is the default 
    behavior for the plotting function used.
  }
\end{figure}

Most of the other settings are the same as before: 
we reuse the alignment parameters (Listing~\ref{lst:create_alignment_params}), 
as well as the experiments (Listing~\ref{lst:create_all_experiments}).
This time, however, we enable twinning (see Subsection~\ref{subsec:twinning}),
by creating twinning parameters.
Creating these parameters is trivial with the default settings.
For clarity, however, we explicitly show the most important arguments:

@@

We run \verb;pirouette; (Listing~\ref{lst:pir_run}) 
and plot the results (Listing~\ref{lst:pir_plot}).
The output is shown in Figure~\ref{fig:example_3}. 
While the error distributions using the best or generative model 
as inference model are very similar, 
the error distributions of the true tree are substantially 
larger than those of the twin tree. 
This is the error made by the mismatch
between the generating species tree model and the tree prior used in inference.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_3/example_3_314/errors.png}
  \caption{
    The inference error made 
    for both a generative tree prior and best candidate model
    compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error.
    Vertical dashed lines show the median error value per distribution.
  }
  \label{fig:example_3}
\end{figure}

>>>>

\richel{

\verb;pirouette; quantifies the influence of a new tree prior on BEAST2's 
inference by measuring the discrepancy between a given/true tree and a posterior 
distribution of phylogenies obtained as result of the inference process. 
Due to stochasticity, posterior trees will generally differ from the given 
phylogeny $\tau_{\mathit{G}}$, even when the tree 
prior and alignment model used for inference are the same as those 
used to generate the alignment.

Measuring this difference allows us to know the baseline error
of the \verb;pirouette; pipeline. We therefore define as 'standard tree priors' 
all the tree priors that are available 
to be used within an inference model (see Table~\ref{tab:options}).

...

[first step]

...

...

@@

@@

} % richel

We show the usage of \verb;pirouette; on an tree generated by a non-standard 
tree model \richel{code at \url{https://github.com/richelbilderbeek/pirouette_example_30}}. 
We use the diversity-dependent tree model \cite{DDD, etienne2011diversity},
which is a birth-death model with a speciation rate that is dependent on the
number of species.

We choose to use a small tree with six taxa, to keep
the calculations short and the figure more readable.
We pick a crown age of ten time units. This value is 
completely arbitrary, but it ties in with the mutation rate 
used in simulating an alignment in the next step.

\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single, 
  label = {lst:create_dd_tree},
  caption = A phylogeny generated by an unknown diversification model.
]
phylogeny <- create_dd_tree(n_taxa = 6, crown_age = 10)
\end{lstlisting}
\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_30/example_30_314/true_tree.png}
  \caption{
    The tree derived from a DD process, 
    as created by listing~\ref{lst:create_dd_tree}.
  }
\end{figure}

We now have all the needed \verb;pirouette; parameters: the alignment 
parameters, the experiments and the error measure parameters.

Next, we enable twinning (see Subsection~\ref{subsec:twinning}),
by creating twinning parameters.
Creating these parameters is trivial with the default settings.
For clarity, however, we explicitly show the most important arguments:

@@

We can finally use the DD tree and \verb;pir_params; to measure the 
inference error made on phylogenies
created by a standard diversification model:

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_30/example_30_314/errors.png}
  \caption{
    The inference error made 
    for both a generative tree prior and best candidate model
    compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error.
    Vertical dashed lines show the median error value per distribution.
  }
  \label{fig:example_30}
\end{figure}

While the error distributions using the best or generative model 
as inference model are very similar, 
the error distributions of the true tree are substantially 
larger than those of the twin tree. 
This is the error made by the mismatch
between the generating species tree model and the tree prior used in inference.
